### 2014-05-21 text/Chapter04.tex
Minor additions to categorical semantics (Chapter 4)

\subsection{Examples of indexed comonads}

@@

\begin{definition}
A \emph{family of comonads} is formed by triples $(\ctyp{\cclrd{r}}{}, \ident{cobind}_{\cclrd{r}}, 
  \ident{counit}_{\cclrd{r}})$ for all $\cclrd{r}$ such that each triple forms a comonad. Given 
$\cclrd{r}, \cclrd{r'}$ such that $\cclrd{r} \leq \cclrd{r'}$, there is also a mapping 
$\iota_{\cclrd{r'}, \cclrd{r}} : \ctyp{\cclrd{r'}}{} \rightarrow \ctyp{\cclrd{r}}{}$ satisfying
certain coherence conditions.
\end{definition}

@@

Indexed comonads provide a semantics to sequential composition, but
additional structure is needed for the semantics of the full coeffect calculus.
Uustalu and Vene~\cite{comonads-notions} additionally require a (\emph{lax semi-}) \emph{monoidal comonad} 
structure, which provides a monoidal operation $\mathsf{m} : C A
\times C B \rightarrow C (A \times B)$ for merging contexts (used in
the semantics of abstraction).

The semantics of the coeffect calculus requires an indexed
lax semi-monoidal structure for combining contexts \emph{as well as} an indexed
\emph{colax} monoidal structure for \emph{splitting} contexts. These are provided
by two families of morphisms (given a coeffect algebra with $\vee$ and $\wedge$):
%
\begin{itemize}
\item $\cmerge_{r, s} : \ccat^r A \times \ccat^s B \rightarrow \ccat^{(r \wedge s)}(A \times B)$ natural
in $A, B$;
\item $\csplit_{r, s} : \ccat^{(r \vee s)}(A \times B) \rightarrow \ccat^r A \times \ccat^s B$ natural
in $A, B$;
\end{itemize}
%
\noindent
The $\cmerge_{r, s}$ operation merges contextual computations with
tags combined by $\wedge$ (greatest lower-bound), elucidating 
the behaviour of $\cmerge_{r, s}$: that merging may result in 
the loss of some parts of the contexts $r$ and $s$.

The $\csplit_{r, s}$ operation splits context-dependent computations
and thus the contextual requirements. To obtain coeffects $r$ and
$s$, the input needs to provide \emph{at least} $r$ and $s$, so the
tags are combined using the $\vee$ (least upper-bound).

For the sake of brevity, we elide the indexed versions of the laws required by Uustalu and Vene
(\eg~most importantly, merging two contexts and then adding the third is
equivalent to merging the last two and then adding the first; similar rule holds is required
for splitting).

>>>>

\paragraph{Examples.}

@@

\begin{definition}
\label{def:flat-family}
A \emph{family of comonads} is formed by triples $(\ctyp{\cclrd{r}}{}, \ident{cobind}_{\cclrd{r}}, 
  \ident{counit}_{\cclrd{r}})$ for all $\cclrd{r}$ such that each triple forms a comonad. Given 
$\cclrd{r}, \cclrd{r'}$ such that $\cclrd{r} \leq \cclrd{r'}$, there is also a mapping 
$\iota_{\cclrd{r'}, \cclrd{r}} : \ctyp{\cclrd{r'}}{} \rightarrow \ctyp{\cclrd{r}}{}$ satisfying
certain coherence conditions.
\end{definition}

@@

Indexed comonads model the semantics of sequential composition, but additional structure is needed
to model the semantics of the flat coeffect calculus. This is where the duality between monads and 
comonads can no longer help us, because context is propagated differently than effects in lambda 
abstraction and application.

Whereas Moggi~\cite{monad-notions} requires \emph{strong} monad to model effectful $\lambda$-calculus, 
Uustalu and Vene~\cite{comonads-notions} require \emph{lax semi-monoidal} comonad to model 
$\lambda$-calculus with contextual properties. The structure requires a monoidal operation:
%
\begin{equation*}
\ident{m} : \ctyp{}{\alpha} \times \ctyp{}{\beta} \rightarrow \ctyp{}{(\alpha \times \beta)}
\end{equation*}
%
The \ident{m} operation is needed in the semantics of lambda abstraction. It represents merging of 
contexts and is used to merge the context of the declaration-site (containing free variables)
and the call-site (containing bound variable). For example, for implicit parameters, this combines
the additional parameters defined in the two contexts.

The semantics of flat coeffect calculus requires operations for \emph{merging}, but also for
\emph{splitting} of contexts. These are provided by \emph{lax} and \emph{oplax} monoidal 
structures. In addition, we also need a lifting operation (similar to $\iota$ from 
Definition~\ref{def:flat-family}) to model sub-coeffecting.

\begin{definition}
Given a flat coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
an \emph{indexed monoidal comonad} is an indexed comonad over the monoid $(\C, \cseq, \cunit)$
equipped with families of operations $\ident{merge}_{\cclrd{r},\cclrd{s}}$, $\ident{split}_{\cclrd{r},\cclrd{s}}$ 
and $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ where:
%
\begin{compactitem}
\item $\ident{merge}_{\cclrd{r},\cclrd{s}}$ is a family of mappings
  $\ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} \rightarrow \ctyp{\cclrd{r}\czip\cclrd{s}}{(\alpha \times \beta)}$
\item $\ident{split}_{\cclrd{r},\cclrd{s}}$ is a family of mappings
  $\ctyp{\cclrd{r}\cpar\cclrd{s}}{(\alpha \times \beta)} \rightarrow \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta}$
\item $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ is a family of mappings
  $\ctyp{\cclrd{r'}}{\alpha} \rightarrow \ctyp{\cclrd{r}}{\alpha}$~ for all $\cclrd{r'}, \cclrd{r}$ such that $\cclrd{r}\,\cleq\,\cclrd{r'}$
\end{compactitem}
%
% We should require lift to be comonad morphism
%
% \tag{\emph{comonad morphism}}
%   \ident{cobind}_{\cclrd{r}, \cclrd{s}}~f \circ \ident{lift}_{\cclrd{r'}\cseq\cclrd{s'}, \cclrd{r}\cseq\cclrd{s}} &=
%   \ident{lift}_{\cclrd{s'}, \cclrd{s}} \circ \ident{cobind}_{\cclrd{r'}, \cclrd{s'}}~(f \circ \ident{lift}_{\cclrd{r'}, \cclrd{r}} )
%
% ... but adapting the coherence conditions for counit/cobind is ugly!
%
\end{definition}

\noindent
The $\ident{merge}_{\cclrd{r},\cclrd{s}}$ operation is the most interesting one. Given two comonadic
values with additional contexts specified by $\cclrd{r}$ and $\cclrd{s}$, it combineds them into a 
single value with additional context $\cclrd{r}\czip\cclrd{s}$. The $\czip$ operation often represents
\emph{greatest lower bound}\footnote{The $\czip$ and $\cpar$ operations are the greatest and least upper 
bounds for the liveness and data-flow examples, but not for implicit parameters. However, they are useful 
as an informal analogy.}, elucidating the fact that merging may result in the loss of some parts of 
the contexts $\cclrd{r}$ and $\cclrd{s}$. We look at examples of this operation in the next section.

The $\ident{split}_{\cclrd{r},\cclrd{s}}$ operation splits a single comonadic value (containing a tuple)
into two separate values. Note that this does not simply duplicate the value, because the additional
context is also split. To obtain coeffects $\cclrd{r}$ and $\cclrd{s}$, the input needs to provide 
\emph{at least} $\cclrd{r}$ and $\cclrd{s}$, so the tags are combined using the $\cpar$, which is often 
the \emph{least upper-bound}\footnotemark[1].

Finally, $\ident{lift}_{\cclrd{r'}, \cclrd{r}}$ is a family of operations that ``forget'' some part of
a context. This models the sub-coeffecting operation and lets us, for example, forget some of the
available implicit parameters, or turn a live context (containing a value) into a dead context (empty).

\paragraph{Simpler.}

\paragraph{Examples.}
